Coherent optical association of a single molecule

### 2020-09-24 paper.tex
Update on Overleaf.

Such ultracold molecular association takes advantage of the much developed cooling and trapping techniques for atoms  as a starting point. To overcome the challenges of small wavefunction overlap and the large release of binding energy of converting atoms to deeply-bound molecules, a  two-step approach has been established to first associate atom pairs into weakly-bound molecules, and then transfer the molecules from this single internal state to a desired rovibrational and electronic state~\cite{Danzl2008, Ni2008,Lang2008, Takekoshi2014, Molony2014, Park2015, Guo2016, Kondov2019, Voges2020}.
So far, all of such association processes  utilized a magnetic Feshbach scattering resonance and have been applied to bialkali molecules. The only exception is Sr$_2$ where narrow linewidth excited states are available and optical association can be driven~\cite{Reinaudi2012,Stellmer2012}. The requirement of a Feshbach resonance to enhance atom-to-molecule wavefunction overlap or the narrow excited state lines limits the generality of the association technique. % to more diverse species.

Here, we demonstrate coherent association of an atom pair to a weakly bound molecule using a two-photon optical Raman transfer via an electronic excited state, without the use of a Feshbach resonance nor a narrow excited state. The resulting single molecule is in a single internal quantum state and predominately in its motional ground state.  Our scheme is based on a theoretically determined vibrational state of the electronic excited state $c^3\Sigma^+$ that has the best transfer Rabi coupling to photon scattering ratio. To further increase this ratio and reduce technical requirements such as intensity stability, we choose an initial and final state where the matrix elements with the excited state are as balanced as possible. This approach minimizes the reliance on system specific properties and can therefore be more generally applied to creating other molecular species or larger molecules atom-by-atom with full quantum state control. 
%(also this result represents full quantum-state control of single [neutral] molecules, the second time ever (first being our NaCs Feshbach molecules))

@@

The challenge of optical Raman transfers of atoms to molecules lies with achieving a high enough Raman Rabi frequency  due to the wavefunction size mismatch with the excited molecular state. While pioneering experiments used weakly bound molecular excited states in the Raman transition to increase the Raman Rabi frequency, such choice of state also increases the scattering during the transfer process to render it incoherent, resulting in molecule loss~\cite{Wynar2000,Rom2004}. 
Therefore, finding the largest Raman Rabi frequency to scattering ratio that enables population transfer with a minimal lost is the first key task.

In a simplified model of a three-level system (as shown in Fig.~\ref{f-theory}), the initial atomic state, an excited intermediate state, and the target weakly-bound molecular state in the ground electronic potential are coupled by two photons, $\Omega_a$ and $\Omega_m$.  %(Ultimately, we calculate with contribution of all vibrational states of the excited electronic potential, $c^3\Sigma^+$)
The Raman Rabi Rate, $ \Omega_a\Omega_m / 2\Delta$ depends on both the matrix element between the atomic state and the excited state, $ \Omega_a  $ and the matrix element between the molecular state and the excited state, $ \Omega_m $. Unlike in atoms, the two matrix elements, $ \Omega_m$ and $ \Omega_a$ are greatly imbalanced. Since $ \Omega_m $ is generally much larger than $ \Omega_a $ due to the wavefunction mismatch between the atomic state and the excited molecular state, the scattering rate predominantly comes from the molecular state, $ \Gamma_e \Omega_m^2 / 2\Delta^2$, where $ \Gamma_e $ is the excited state linewidth and $ \Delta $ is the single photon detuning. We choose the power of the Raman beams to be equal (which gives the largest Raman Rabi rate), so there is an additional factor of 2 in the scattering rate, since the molecular state scatters off of both Raman beams at roughly the same single photon detuning. Thus, the ratio between the Raman Rabi frequency and the scattering rate, $ \Omega_a/\Omega_m \times \Delta/\Gamma_e $, depends on the ratio of the two matrix elements and how far detuned the laser is from the transition in units of the linewidth. To ensure a coherent process, larger detuning is preferred, but if it's too large, the Raman Rabi rate may become too small. With the multiple excited vibrational states present in a molecular potential, the total scattering rates becomes a sum of the scattering rates over all the states. Since for a typical molecular potential the spacing between states
decreases with increasing vibrational levels,
a weakly bound excited molecular state limits the maximum usable detuning and therefore the effectiveness of the scattering reduction by increasing the detuning for the Raman transition. Furthermore, due to similarity in wavefunction size, the weakly bound molecular ground state has a strong coupling with the atomic excited state which contributes significantly to the total scattering rate. The rate of such a scattering process is larger for weakly bound molecular excited states, and is proportional to $1/\delta_{thresh}^2$,
where $\delta_{thresh}$ is the detuning from the dissociation threshold. (Should we also mention scattering of the atomic initial state too? This should affect the Raman process as well)
% This further increases the scattering rate for a weakly bond excited state.
Based on these considerations, % a weakly bound excited molecular state may not be the best choice for the intermediate state in the Raman transfer. Thus, 
we calculated both the Raman Rabi frequency and the scattering rate at different detunings (full result is shown in SM). As seen in figure (1....), we found a distinct regime from what have been explored previously, namely, despite a smaller Raman Rabi frequency at deeply bound intermediate states, the Rabi frequency to scattering ratio is much higher as compared to a weakly bound intermediate state. As a result, we use a deeply bound molecular excited state ($v'=0$ of $c^3\Sigma^+$) to drive the Raman transition. (mention the trend, also v=12 will work as similarly as v=0)

@@

In addition to the intermediate state, the choice of the initial and final states are critical for both fundamental and technical reasons. As previously discussed, the ratio of the Raman Rabi rate to the scattering rate is proportional to the $ \Omega_a/\Omega_m $ ratio. The smaller this ratio is, the farther the laser needs to be detuned to achieve the same Raman Rabi rate to scattering rate ratio. This ratio can be made smaller by increasing the coupling of the atomic state with the excited molecular state. Due to the small size of the molecular wavefunction, the coupling between the ground atomic state and the excited molecular state is approximately proportional to the value of the relative atomic wavefunction at short distance within the molecular potential. In additional to the confinement, this value is related to the interaction between the two atoms. For states with a large scattering length (positive or negative), the phase shift in the relative wavefunction between the atoms can significantly increase the short range wavefunction (add figure?). The increase in the coupling is proportional to (quote/cite Olive's equation?). For our system, among the stable spin combinations, 4422 and 3311 both have small scattering lengths of $\abs{a_{4422}}= 30.4a_0, \abs{a_{3311}} = 13.7a_0 $ respectively and 3322 has a large and negative scattering length of $ \abs{a_{3322}} = 693.8a_0 $ (interaction shift $\approx$ binding?)~\cite{Hood2019}. In addition to the increased atomic coupling, $ \Omega_a $, with the 3322 hyperfine combination, coupled channel calculations show that the most weakly bound molecular state that is predominantly this spin combination also has reduced coupling, $ \Omega_m $, with the excited state when compared with the 4422 and 3311 bound states. Therefore, using an initial 3322 hyperfine combination results in a $ \Omega_a/\Omega_m$ ratio of about 0.05 instead of a ratio of about 0.003 with the 4422 or 3311 combinations. In addition to reducing how far the laser needs to be detuned from the transition, a larger $ \Omega_a/\Omega_m$ ratio also relaxes the intensity stability requirement. The position of the resonance depends on the laser power predominantly through the AC Stark shift on the molecular state, $ \Omega_m^2 / 2\Delta $. The ratio of the AC Stark shift to the Raman Rabi frequency is $ \Omega_m / \Omega_a $, Thus, the laser needs to be stabilized to better than the inverse of this ratio, $ \Omega_a/\Omega_m $, to fluctuate by less than a linewidth. Thus, instead of a required intensity stability of $0.3 \% $ for the 4422 combination, the 3322 combination only requires the intensity to be stabilized to $ 5\% $.  Thus, we use the 3322 spin combination as our initial state and drive to the first bound state for the 3322 spin combination. (We might want to mention that those ratios are for 80 kHz spherical trap if we want to give more details about the coupled channel calculation anywhere...)

>>>>

Such ultracold molecular association takes advantage of the much developed cooling and trapping techniques for atoms as a starting point. To overcome the challenges of small wavefunction overlap and the large release of binding energy of converting atoms to deeply-bound molecules, a  two-step approach has been established to first associate atom pairs into weakly-bound molecules, and then transfer the molecules from this single internal state to a desired rovibrational and electronic state~\cite{Danzl2008, Ni2008,Lang2008, Takekoshi2014, Molony2014, Park2015, Guo2016, Kondov2019, Voges2020}.
So far, all of such association processes  utilized a magnetic Feshbach scattering resonance and have been applied to bialkali molecules. The only exception is Sr$_2$ where narrow linewidth excited states are available and optical association can be driven coherently~\cite{Reinaudi2012,Stellmer2012}. The requirement of a Feshbach resonance to enhance atom-to-molecule wavefunction overlap or the existence of narrow excited state lines limits the generality of the association technique. % to more diverse species.

Here, we demonstrate coherent association of an atom pair to a weakly bound molecule using a two-photon optical Raman transfer via an electronic excited state, without the use of a Feshbach resonance nor a narrow excited state, schematically shown in Fig.~\ref{f-theory}. The resulting single molecule is in a single internal quantum state and predominately in its motional ground state.  Our scheme is based on using a theoretically determined vibrational state of the electronic excited state $c^3\Sigma^+$ that has the best transfer Rabi coupling to photon scattering ratio. To further increase this ratio and reduce technical requirements such as intensity stability, we choose an initial and final state where the matrix elements with the excited state are as balanced as possible. This approach minimizes the reliance on system specific properties and could therefore be applied to creating other molecular species or larger molecules atom-by-atom with full quantum state control. 
%(also this result represents full quantum-state control of single [neutral] molecules, the second time ever (first being our NaCs Feshbach molecules))

@@

The challenge of optical Raman association of molecules from atoms lies with achieving a high enough Raman Rabi frequency despite the wavefunction size mismatch between the atomic state and the excited molecular state. While pioneering experiments used weakly bound molecular excited states in the Raman transition to increase the Raman Rabi frequency, such choice of state also increases the scattering during the transfer process to render it incoherent, resulting in molecule loss~\cite{Wynar2000,Rom2004}. 
Therefore, finding the largest Raman Rabi frequency to scattering ratio that enables population transfer with minimal loss is the first key task.

In a simplified model of a three-level system (as shown in Fig.~\ref{f-theory}), we consider the initial atomic state, an excited intermediate state, and the target weakly-bound molecular state in the ground electronic potential. The initial atomic state and target weakly-bound state are coupled by two photons, $\Omega_a$ and $\Omega_m$ to the excited intermediate state, with single photon detuning $ \Delta $, and two photon detuning, $ \delta$.  %(Ultimately, we calculate with contribution of all vibrational states of the excited electronic potential, $c^3\Sigma^+$)
The Raman Rabi Rate for a detuned transfer, $ \Omega_a\Omega_m / 2\Delta$ depends on both of these matrix elements and the single photon detuning. Unlike for Raman transitions in atoms, the two matrix elements, $ \Omega_m$ and $ \Omega_a$ are greatly imbalanced. Since $ \Omega_m $ is generally much larger than $ \Omega_a $ due to the wavefunction mismatch between the atomic state and the excited molecular state, the scattering rate predominantly comes from the molecular state, $ \Gamma_e \Omega_m^2 / 2\Delta^2$, where $ \Gamma_e $ is the excited state linewidth. We choose the power of the Raman beams to be equal (which gives the largest Raman Rabi rate at a fixed total power), so there is an additional factor of 2 in the scattering rate, since the molecular state scatters off of both Raman beams at roughly the same single photon detuning. Thus, the ratio between the Raman Rabi frequency and the scattering rate, $ \Omega_a/\Omega_m \times \Delta/\Gamma_e $, depends on the ratio of the two matrix elements and how far detuned the laser is from the transition in units of the linewidth. To ensure a coherent process, larger detuning is preferred.

With the multiple excited vibrational states present in a molecular potential, the total scattering rates becomes a sum of the scattering rates over all the states, and there are many possible intermediate states to choose from. To obtain the full picture, we perform a full calculation of the Raman Rabi frequency and scattering rate at different detunings (full result in SM). As shown in figure ~\ref{f-theory}, the ratio of the Raman Rabi rate to scattering rate can be made better for more deeply bound states compared to weakly bound states. For a weakly bound molecular state, the spacing between states is generally smaller compared to deeply bound states and thus %Since for a typical molecular potential the spacing between states
%decreases with increasing vibrational levels,
%a weakly bound excited molecular state 
limits the maximum usable detuning and therefore the effectiveness of the scattering reduction by increasing the detuning. Furthermore, due to similarity in wavefunction size, the weakly bound molecular ground state has strong coupling with the atomic excited state which contributes significantly to the total scattering rate. The rate of such a scattering process is proportional to $1/\delta_{thresh}^2$,
where $\delta_{thresh}$ is the detuning from the dissociation threshold. (Should we also mention scattering of the atomic initial state too? This should affect the Raman process as well)
% This further increases the scattering rate for a weakly bond excited state.
Based on these considerations, % a weakly bound excited molecular state may not be the best choice for the intermediate state in the Raman transfer. Thus, 
% we calculated both the Raman Rabi frequency and the scattering rate at different detunings (full result is shown in SM). As seen in figure (1....), 
we found a distinct optimal regime from what has been explored experimentally previously. Namely, despite a smaller Raman Rabi frequency at deeply bound intermediate states, the Rabi frequency to scattering rate ratio is much higher as compared to a weakly bound intermediate state. As a result, we use a deeply bound molecular excited state ($v'=0$ of $c^3\Sigma^+$) to drive the Raman transition. (mention the trend, also v=12 will work as similarly as v=0)

@@

In addition to the intermediate state, the choice of the initial and final states is critical. As previously discussed, the ratio of the Raman Rabi rate to the scattering rate is proportional to the $ \Omega_a/\Omega_m $ ratio. The smaller this ratio is, the farther the laser needs to be detuned to achieve the same Raman Rabi rate to scattering rate ratio, which lowers the overall Raman Rabi Rate. This ratio can be made smaller by increasing the coupling of the atomic state with the excited molecular state. Due to the small size of the molecular wavefunction, the coupling between the ground atomic state and the excited molecular state is approximately proportional to the value of the relative atomic wavefunction at short distance within the molecular potential. In addition to the confinement, this value is related to the interaction between the two atoms. For states with a large scattering length (positive or negative), the phase shift in the relative wavefunction between the atoms can significantly increase the short range wavefunction (add figure?). The increase in the coupling is proportional to (quote/cite Olive's equation?). For our system, among the stable spin combinations, 4422 and 3311 both have small scattering lengths of $\abs{a_{4422}}= 30.4a_0, \abs{a_{3311}} = 13.7a_0 $ respectively and 3322 has a large and negative scattering length of $ \abs{a_{3322}} = 693.8a_0 $ (interaction shift $\approx$ binding?)~\cite{Hood2019}. In addition to the increased atomic coupling, $ \Omega_a $, with the 3322 hyperfine combination, coupled channel calculations show that the most weakly bound molecular state that is predominantly this spin combination also has reduced coupling, $ \Omega_m $, with the excited state when compared with the 4422 and 3311 bound states. Therefore, using an initial 3322 hyperfine combination results in a $ \Omega_a/\Omega_m$ ratio of about 0.05 instead of a ratio of about 0.003 with the 4422 or 3311 combinations. In addition to reducing how far the laser needs to be detuned from the transition, a larger $ \Omega_a/\Omega_m$ ratio also relaxes the intensity stability requirement, a key potential technical limitation. The position of the resonance depends on the laser power predominantly through the AC Stark shift on the molecular state, $ \Omega_m^2 / 2\Delta $. The ratio of the AC Stark shift to the Raman Rabi frequency is $ \Omega_m / \Omega_a $, Thus, the laser needs to be stabilized to better than the inverse of this ratio to fluctuate by less than a linewidth. Thus, instead of a required intensity stability of $0.3 \% $ for the 4422 combination, the 3322 combination only requires the intensity to be stabilized to $ 5\% $.  Thus, we use the 3322 spin combination as our initial state and drive to the first bound state for the 3322 spin combination. (We might want to mention that those ratios are for 80 kHz spherical trap if we want to give more details about the coupled channel calculation anywhere...)
